Summer 2025 ToDo

### 2025-09-19 Relations/Paper/Well-foundedness.tex
Well-foundedness update

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in many cases of interests.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
For example, the file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$ --- which is clearly incompatible with effective semantics.
\apcomment{Make sure the file exists, recover it if necessary!}

@@

\begin{figure}[ht]
\centering  
\begin{tikzpicture}[
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

@@

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.

To explore how these definitions interrelate it helps to consider the following corresponding weaker definitions (which are formalized in \texttt{WFWeakDefinitions.agda}). 
The weaker definitions help us to explore how the standard well-founded definitions interrelate. This is made clear in the following diagram.

\subsection{New version: Synopsis and Outline} 
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the
implication from sequential to inductive wellfoundedness
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {$\accCor$} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  (In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.)
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

>>>>

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in many cases of interests.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
For example, the file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$ --- which is clearly incompatible with effective semantics.
\apcomment{Make sure the file exists (Done), check it wrt other new stuff in the FPLA repo.}

@@

\begin{figure}[hb]
\centering  
\begin{tikzpicture}[
  scale=0.7,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

@@

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.
\apcomment{Move this comment elsewhere? Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

The logical relationships between the various definitions of well-foundedness are summarized in 
Figure \ref{fig:WF}.
Again, all of the definitions appearing in this figure are classically equivalent.
But constructively, some implications are not provable in general.
Formalizing these implications enabled us to identify where exactly classical logic 
is used, and how much of it is truly needed.
An arrow from one node to another with no label indicates direct logical implication,
provable constructively with no additional assumptions. An arrow with a label indicates an additional 
(semi-)classical hypothesis needed to establish the implication. All of these facts have been 
formalized within the WellFounded subdirectory of the Agda repository.

\apcomment{Things to do:}
\begin{itemize}
  \item Reiterate/articulate the role of classical principles in establishing relationships between definitions of 
    well-foundedness constructively 
  \item List/discuss? classical principles appearing in the figure 
  \item Discuss the figure overall -- refer to the enumerated list below 
  \item Create ``narrative bridges'' between these items 
  \item Final thing: Instantiate some of the classical principles on simple examples of term rewrite systems.
\end{itemize}

% As the figure illustrates, the concept of well-foundness is very rich from the vantage point of constructive logic.
% Our analysis resulted in the following principles, 
%
% The additional classical hypotheses needed to recover some of the implications on the diagram 
%
% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.  
%
% The weaker definitions help us to explore how the standard well-founded definitions interrelate. This is made clear in the following diagram.
%
\subsection{New version: Synopsis and Outline} 
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the
implication from sequential to inductive wellfoundedness
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {$\accCor$} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  (In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.)
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.
